---
title: "User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## The problem

An uncaught error inside a plain `shiny::observe()` or `shiny::observeEvent()` terminates the
whole Shiny session for that user. This is easy to miss, because the *other* common case - an
error inside a `reactive()` that's consumed only by a render output - behaves completely
differently and does **not** crash the session.

Shiny's render functions (`renderPlot()`, `renderText()`, etc.) catch an error raised in their
own body and convert it into the output's *value*: it gets serialized as
`<div class="shiny-output-error">...</div>` and sent to the browser exactly like a successful
render would be. From the reactive flush loop's point of view, that output "succeeded" - it
produced a value, just an error-flagged one. Nothing propagates further, and every other
output/observer in the session keeps working normally afterward.

A bare `observe()`/`observeEvent()` has no such slot to catch into - its whole purpose is a side
effect, not a value, so there's no DOM element to localize the error to. An uncaught error here
keeps unwinding through Shiny's generic flush-loop logging, with no per-observer recovery path.
In practice this can leave a session in a state where the browser's own connection indicator
still shows "connected," but nothing updates again without a reload - indistinguishable from a
crash as far as the user is concerned.

There's a third, separate mechanism worth knowing about: `shiny::req()` and `shiny::validate()`
deliberately throw a specially classed `shiny.silent.error` condition that Shiny's reactive
machinery recognizes *everywhere* (outputs and observers alike) as "stop evaluating silently -
this isn't a bug, don't show anything, don't log." That's intentional control flow, not an error
to report - and it's why a naive `tryCatch(error = ...)` wrapper around an observer body would be
wrong: it would also catch `req()`/`validate()`'s silent stops and misreport every one of them as
a real failure.

## SafeObserve() / SafeObserveEvent()

`SafeObserve()` and `SafeObserveEvent()` are drop-in replacements for `shiny::observe()` and
`shiny::observeEvent()` that catch genuine errors - and only genuine errors, correctly passing
`shiny.silent.error` through unchanged - so a bug in one observer degrades gracefully instead of
taking down the whole session.

```{r}
library(shiny)
library(SafeShiny)

# Before: a bug here disconnects every user in this session.
server <- function(input, output, session) {
  observeEvent(input$go, {
    result <- 1 / input$denominator # errors if denominator is non-numeric/zero in some paths
    showNotification(paste("Result:", result))
  })
}

# After: the same bug just shows a notification and logs the error - the session survives.
server <- function(input, output, session) {
  SafeObserveEvent(input$go, {
    result <- 1 / input$denominator
    showNotification(paste("Result:", result))
  }, onError = function(e) {
    showNotification(paste("Something went wrong:", conditionMessage(e)), type = "error")
  })
}
```

Both functions accept every argument their `shiny` counterparts do (`ignoreInit`, `once`,
`priority`, `domain`, ...), plus:

- `onError` - a function called with the caught condition, for application-specific handling.
- `quiet` - set `TRUE` to suppress the default `message()` logged when an error is caught (you
  still get `onError`, if supplied).
- `trackTime`/`label` - see the next section.

`req()`/`validate()` keep working exactly as before - `SafeObserveEvent(input$go, { req(input$x); ... })`
behaves identically to the plain `shiny::observeEvent()` version when `input$x` isn't ready yet.

## Execution-time tracking

A long-standing, hard-to-answer question when an app feels slow is: *how much of that time is
actually spent in my own business logic, versus Shiny's own reactive-graph bookkeeping
(invalidation, scheduling, flushing)?* Profiling tools like `Rprof()` or `reactlog` can show call
counts and stack traces, but splitting "my code" from "Shiny's machinery" cleanly from that
output is its own, separate piece of work - exactly the kind of split a team investigating a
slow file-upload path might want to make but not have time to finish measuring directly.

Passing `trackTime = TRUE` to `SafeObserve()`/`SafeObserveEvent()` times the wrapped expression
(whether it finishes normally, is caught by `onError`, or hits a `req()`/`validate()` silent stop
- all three consume real wall-clock time) and records it, keyed to the current Shiny session so
concurrent users' timings never mix.

`trackTime` defaults to `getOption("SafeShiny.trackTime", FALSE)`, so setting
`options(SafeShiny.trackTime = TRUE)` once (before the wrappers are created) turns tracking on for
every `Safe*` call that doesn't pass `trackTime` itself - including ones created by other
packages such as generated observers and reactives. An explicit `trackTime = FALSE` still wins.

```{r}
server <- function(input, output, session) {
  SafeObserveEvent(input$upload, {
    processUploadedFile(input$upload$datapath)
  }, trackTime = TRUE, label = "process_upload")

  SafeObserveEvent(input$recompute, {
    recomputeDownstreamValues()
  }, trackTime = TRUE, label = "recompute")

  observeEvent(input$show_timing, {
    print(SummarizeSafeShinyTiming(session = session))
  })
}
```

`GetSafeShinyTiming()` returns a per-label summary (call count, total/mean/last elapsed time).
`SummarizeSafeShinyTiming()` goes one step further and compares the sum of everything tracked
against the wall-clock time elapsed since the first tracked call in the session:

```{r}
summ <- SummarizeSafeShinyTiming(session = session)
print(summ)
#> SafeShiny timing summary
#>   Tracked user-code time: 2.341s across 2 label(s)
#>   Wall-clock elapsed since first tracked call: 3.102s
#>   Approx. untracked time (Shiny overhead + anything not wrapped): 0.761s
#>   NOTE: 'untracked time' only approximates Shiny's own reactive-graph overhead once
#>         EVERY relevant observer in this session is wrapped with trackTime = TRUE -
#>         otherwise it also includes time spent in anything left unwrapped.
```

**The caveat in that last line matters.** `untracked_time` is `wall_clock_elapsed -
total_tracked_time` - it is only a meaningful estimate of genuine Shiny reactive-graph overhead
once *every* relevant observer/reactive in the session's critical path is wrapped with
`trackTime = TRUE`. If some business logic runs inside an un-wrapped `observeEvent()` or
`reactive()`, its time ends up lumped into `untracked_time` too, and you'll overestimate how much
of the gap is really Shiny's own bookkeeping rather than just unmeasured application code.

Use `GetSafeShinyTimingRaw()` if you need the per-call detail (including whether each call
finished `"ok"`, hit a `"silent"` `req()`/`validate()` stop, or was caught as an `"error"`) rather
than the aggregated-by-label view `GetSafeShinyTiming()` returns. Call `ResetSafeShinyTiming()` to
clear a session's recorded timings (e.g. between two different scenarios you want to compare
independently).

`PlotSafeShinyFlame()` draws the same records as a nested flame chart (time on the x-axis, nesting
depth on the y-axis): a tracked call evaluated synchronously inside another one - for instance a
`SafeReactive()` read from a `SafeObserve()` - is drawn directly above its parent. Nesting is
inferred from a per-session stack of running tracked calls, so it is only reliable for
synchronous code (`promises`/`future` intervals interleave).

## SafeRender() / SafeRenderPlot() / SafeRenderUI() / SafeRenderTable() / SafeDownloadHandler()

These are a **different shape of wrapper** from `SafeObserve()`/`SafeObserveEvent()`, and it's
worth being precise about why. A `reactive()`/render body consumed by an output, and a
`shiny::downloadHandler()`'s `content` function, both already degrade gracefully on their own -
confirmed empirically (minimal apps driven through a real session via `chromote`):

- A `reactive()` feeding `renderPlot()`/`renderText()`/etc.: the session stays fully alive, Shiny
  shows its own local error display for that one output.
- `shiny::downloadHandler()`: a `content` error produces an HTTP 500 for that one download
  request; the session stays fully alive afterward (a separate, unrelated observer kept firing
  normally).

So none of `SafeRender()`/`SafeRenderPlot()`/`SafeRenderUI()`/`SafeRenderTable()`/
`SafeDownloadHandler()` exist to *prevent a crash* - nothing was going to crash. What they give
you is a hook into an error Shiny is already silently containing, so the *application* can react
to it too (log it, show a notification, roll it into some other piece of application state) -
something native Shiny gives you no way to do. Accordingly, unlike `SafeObserve()`/
`SafeObserveEvent()` (which *swallow* a genuine error, because nothing else would), these five
**catch, call `onError()`, then re-raise the original error unchanged** - Shiny's own native
degradation (the red error box, the HTTP 500) still happens exactly as it does today. The
`shiny.silent.error` passthrough is identical to `SafeObserve()`/`SafeObserveEvent()`: a
`req()`/`validate()` silent stop is always re-raised unchanged and never reaches `onError()`.

```{r}
library(shiny)
library(SafeShiny)

server <- function(input, output, session) {
  output$plot <- SafeRenderPlot({
    plotMyData(input$compound)
  }, onError = function(e) {
    # Shiny already shows its own error box for this plot - this hook is extra, e.g. to flag
    # the failure in some other piece of application state.
    recordPlotFailure(input$compound, conditionMessage(e))
  })
}
```

`SafeRenderPlot()`/`SafeRenderUI()` are thin wrappers over the generic `SafeRender(renderFunc,
expr, ...)`, which works with any Shiny-style render-function generator - so a custom or
third-party one can be wrapped the same way: `SafeRender(my_pkg::renderWidget, { ... })`.

### SafeRenderTable() and DT - read this before relying on it

The MMV apps this package was built for use `DT::renderDT()` (the current name for what used to
be `DT::renderDataTable()` - same function), not base `shiny::renderTable()`, so that's what
`SafeRenderTable()` wraps. Checked empirically before assuming it behaves like the base render
functions above - **it doesn't, quite**. A `DT::renderDT()` body made to throw: the session does
survive the same way (a second, unrelated observer kept firing normally afterward) - but unlike
`renderPlot()`/`renderText()`, which show a visible `"An error has occurred"` box, a failed
`DT::renderDT()` renders as a **silently blank, hidden widget** - no `shiny-output-error` class,
no message, nothing in the DOM indicating anything went wrong.

So for DT tables specifically, `onError` isn't just an observability nice-to-have layered on top
of a working native display - it is, practically, **the only way the user or the app finds out
the table failed at all**. Strongly consider always pairing `SafeRenderTable()` with an `onError`
that shows a `shiny::showNotification()` or similar, since DT itself won't tell anyone:

```{r}
output$table <- SafeRenderTable({
  myData()
}, onError = function(e) {
  showNotification("This table failed to load - see the server log.", type = "error")
})
```

### SafeDownloadHandler() - a real limitation, not just a caveat

`SafeDownloadHandler()` wraps `shiny::downloadHandler()`'s `content` function as a whole. It
gives you **one** failure hook for "the whole download failed" - it does **not** give per-item
resilience inside a loop that builds several pieces of one download. The motivating real-world
case: a multi-sheet Excel report where one compound's sheet fails to generate shouldn't abort the
*other* compounds' sheets in the same workbook. `SafeDownloadHandler()` alone does not fix that -
if one iteration of such a loop throws, the whole `content` call still fails, same as it would
without this wrapper. Making the other iterations survive needs a `tryCatch` placed at *each loop
iteration*, inside the report-generation code itself - that's specific to that code's structure
and is out of scope for a generic wrapper like this one.

```{r}
output$report <- SafeDownloadHandler(
  filename = "report.xlsx",
  content = function(file) {
    # SafeDownloadHandler() catches a failure in this whole function. It does NOT, by itself,
    # stop one bad compound's sheet from aborting the rest - that needs its own tryCatch per
    # iteration, inside writeReportWorkbook() itself.
    writeReportWorkbook(file)
  },
  onError = function(e) message("report generation failed: ", conditionMessage(e))
)
```

## SafeReactive()

`SafeReactive()` wraps `shiny::reactive()` and belongs to the observability family above, not the
swallowing one: it **catches, calls `onError()`, then re-raises the original error unchanged**, so
whatever consumes the reactive still gets Shiny's own native degradation. It exists for two
reasons that have nothing to do with crash prevention:

1. **A hook for app-level bookkeeping.** A failing computation can be rolled into your own status
   tracking (e.g. so a report-generation step knows to skip that item), which Shiny's local error
   box cannot do.
2. **Timing coverage of the computation layer.** In a typical app the real compute time lives in
   plain `reactive()` expressions, not in observers (which mostly dispatch) or renders (which
   mostly format). `trackTime = TRUE` on `SafeReactive()` measures that layer, making the
   `untracked_time` estimate in `SummarizeSafeShinyTiming()` far more meaningful.

```{r}
server <- function(input, output, session) {
  fit <- SafeReactive({
    expensiveFit(input$n)
  }, onError = function(e) recordFitFailure(conditionMessage(e)),
     trackTime = TRUE, label = "fit")

  output$summary <- renderPrint(fit())
}
```

Normal reactive caching is unchanged: once the expression has thrown, reading the reactive again
before its next invalidation re-raises the same cached error without recomputing, so `onError()`
fires once per failed computation, not once per read. `req()`/`validate()` silent stops are
re-raised unchanged and never reach `onError()`.

## Roadmap

Implemented: `SafeObserve()`/`SafeObserveEvent()` (swallowing - the only place nothing else would
contain the error), and `SafeReactive()`, `SafeRender()`/`SafeRenderPlot()`/`SafeRenderUI()`/
`SafeRenderTable()`/`SafeDownloadHandler()` (observability-only - Shiny already contains these
on its own).
